---
title: "Movielens Recommendation System"
output:
  pdf_document: default
  html_document: default
date: "12.01.2022"
---

### **1. Introduction**
The movielens dataset consists of 9000055 observations of six variables. It includes \

- **userId** - the user identification number \
- **movieId** - the identification number of each movie \
- **timestamp** - the date and time of the rating \
- **title** - the movie title \
- **genres** - the movie genre \
- **rating** - the movie rating raging from 0-5 in 0.5 steps \

The aim is to predict the rating as an outcome using the columns aboth.

### **2. Analysis**

#### **2.1 The code provided to generate edx dataset and data format modification**\
First I used the code provided to generate the edx data set. I changed the data formats and used the ["POSIXct"](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/as.POSIX*)-function to generate a column for year, month and weekday from the $timestamp column. 
```{r, warning = FALSE, echo = FALSE, results = 'hide', message = FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
library(tidyverse)
library(caret)
library(data.table)
dl <- tempfile()
options(timeout=500)
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))
movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)
rm(dl, ratings, movies, test_index, temp, movielens, removed)

edx$timestamp <- as.POSIXct(edx$timestamp, origin = "1970-01-01")
edx$weekday_full_name <- weekdays(edx$timestamp)
edx$month <- format(edx$timestamp, format = "%m")
edx$year <- format(edx$timestamp, format= "%Y")
edx$month <- as.numeric(edx$month)
edx$year <- as.numeric(edx$year)
edx$weekday <- recode_factor(edx$weekday_full_name, 
                             'Monday'="1",
                             'Tuesday'="2",
                             'Wednesday'="3",
                             'Thursday'="4",
                             'Friday'="5",
                             'Saturday'="6",
                             'Sunday'="7")
edx$weekday <- as.numeric(edx$weekday)
edx$genres <- as.factor(edx$genres)
edx$userId <- as.numeric(edx$userId)

validation$timestamp <- as.POSIXct(validation$timestamp, origin = "1970-01-01")
validation$year <- format(validation$timestamp, format= "%Y")
validation$year <- as.numeric(validation$year)
validation$weekday_full_name <- weekdays(validation$timestamp)
validation$month <- format(validation$timestamp, format = "%m")
validation$month <- as.numeric(validation$month)
validation$weekday <- recode_factor(validation$weekday_full_name, 
                             'Monday'="1",
                             'Tuesday'="2",
                             'Wednesday'="3",
                             'Thursday'="4",
                             'Friday'="5",
                             'Saturday'="6",
                             'Sunday'="7")
validation$weekday <- as.numeric(validation$weekday)
```

#### **2.2 Split the data into train and test set** \
Using to caret function 'createDataPartition' the edx data set is splitted into a test set (20%) and a training set (80%).

```{r, warning = FALSE, echo = TRUE, results = 'hide', message=FALSE}
data <- edx %>% select(rating, weekday, month, year, genres, movieId, userId)
test_index <- createDataPartition(data$rating, times = 1, p = 0.2, list = FALSE)
train_set <- data %>% slice( - test_index)
test_set <- data %>% slice(test_index)
```

#### **2.3 Calculate baseline RMSE** \
\
The root mean square error (RMSE) is significantly higher than 1. The objective is to create an algorithm that achieves a RMSE lower than 1.06071.

```{r,  warning = FALSE, echo = FALSE, results = 'hide', message = FALSE}
baseline_mean <- mean(train_set$rating)
baseline_RMSE <- RMSE(test_set$rating, baseline_mean)
baseline_RMSE
if(!require("knitr")) install.packages("knitr", repos = "http://cran.us.r-project.org")
library("knitr")
```

```{r, echo = FALSE}
rmse_results <- data.frame(Method = "baseline RMSE", RMSE = "1.06071")
kable(rmse_results)
```




#### **2.4 Create a prediction model** \
\
Create a prediction model using the \
- 'user-effect' \
- 'movie-effect' \
- 'genres-effect' \
- 'year-effect' \
- 'month-effect' \
- 'weekday-effect' \

#### **2.4.1 The effect of user**\
There were a total of 69878 different users that rated the movies.
```{r, echo = FALSE, results = 'hide'}
n_distinct(edx$userId)
```

````{r,  warning = FALSE, echo = TRUE, results = 'hide'}
userId_average <- train_set %>%
  group_by(userId) %>%
  summarize(userId_mean = mean(rating - baseline_mean))

ratings_prediction <- test_set %>%
  left_join(userId_average, by = "userId") %>%
  mutate(prediction = baseline_mean + userId_mean) %>%
  pull(prediction)

RMSE_user_effect <- RMSE(ratings_prediction, test_set$rating, na.rm=T)
````

````{r, echo = FALSE}
rmse_results <- rmse_results %>% rbind(c("effect of user", round(RMSE_user_effect, 6)))
kable(rmse_results)
````


#### **2.4.2 The effect of user and movie**\
There were a total of 10677 different movies rates.
``` {r, echo = FALSE, results = 'hide'}
n_distinct(edx$movieId)
```

````{r,  warning = FALSE, echo = TRUE, results = 'hide', messages = F}
movieId_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  group_by(movieId) %>%
  summarize(movieId_mean = mean(rating - baseline_mean - userId_mean))

ratings_prediction <- test_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  mutate(prediction = baseline_mean + userId_mean + movieId_mean) %>%
  pull(prediction)

RMSE_user_movie_effect <- RMSE(ratings_prediction, test_set$rating, na.rm = T)
````

````{r, echo = FALSE, results = 'hide'}
rmse_results <- rmse_results %>% rbind(c("effect of user and movie", round(RMSE_user_movie_effect, 6)))
kable(rmse_results)
````

#### **2.4.3 The effect of user, movie and genres**\
There were a total of 797 different genres.
```{r, echo = FALSE, results = 'hide'}
n_distinct(edx$genres)
```

````{r,  warning = FALSE, echo = TRUE, results = 'hide', messages = FALSE}
genres_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  group_by(genres) %>%
  summarize(genre_mean = mean(rating - baseline_mean - userId_mean - movieId_mean))

ratings_prediction <- test_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by = "genres") %>%
  mutate(prediction = baseline_mean + userId_mean + movieId_mean + genre_mean) %>%
  pull(prediction)

RMSE_user_movie_genre_effect <- RMSE(ratings_prediction, test_set$rating, na.rm=T)
````

````{r, echo = FALSE}
rmse_results <- rmse_results %>% rbind(c("effect of user, movie and genre", round(RMSE_user_movie_genre_effect, 6)))
kable(rmse_results)
````

#### **2.4.4 The effect of user, movie, genres and year**\
Rating frequency was different over the years and there was a drop in ratings in 1998. The mean rating decreased over the years from 4.0 in 1995 to 3.46 in 2009.
````{r, warning = FALSE, echo = TRUE, results = 'hide', messages = FALSE}
years_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by = "genres") %>%
  group_by(year) %>%
  summarize(year_mean = mean(rating - baseline_mean - userId_mean - movieId_mean - genre_mean))

ratings_prediction <- test_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by = "genres") %>%
  left_join(years_average, by = "year") %>%
  mutate(prediction = baseline_mean + userId_mean + movieId_mean + genre_mean + year_mean) %>%
  pull(prediction)

RMSE_user_movie_genre_year_effect <- RMSE(ratings_prediction, test_set$rating, na.rm = T)
````

````{r, echo = FALSE}
rmse_results <- rmse_results %>% rbind(c("effect of user, movie, genre and year", round(RMSE_user_movie_genre_year_effect, 6)))
kable(rmse_results)
````

#### **2.4.5 The effect of user, movie, genres, year and month**\
Users gave more ratings in October, November and December, and ratings tended to be better in those months.
````{r,  warning = FALSE, echo = TRUE, results = 'hide', messages = FALSE}
month_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by= "genres") %>%
  left_join(years_average, by = "year") %>%
  group_by(month) %>%
  summarize(month_mean = mean(rating - baseline_mean - userId_mean - movieId_mean - genre_mean - year_mean))

ratings_prediction <- test_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by = "genres") %>%
  left_join(years_average, by = "year") %>%
  left_join(month_average, by = "month") %>%
  mutate(prediction = baseline_mean + userId_mean + movieId_mean + genre_mean + year_mean + month_mean) %>%
  pull(prediction)

RMSE_user_movie_genre_year_month_effect <- RMSE(ratings_prediction, test_set$rating, na.rm = T)
`````

````{r, echo = FALSE}
rmse_results <- rmse_results %>% rbind(c("effect of user, movie, genre, year and month", round(RMSE_user_movie_genre_year_month_effect, 6)))
kable(rmse_results)
````

#### **2.4.6 The effect of user, movie, genres, year, month and weekday**\
Users gave more ratings in the beginning of the week than on the weekend.The mean rating tended to be slightly worse on Wednesdays and Thursdays.
````{r,  warning = FALSE, echo = TRUE, results = 'hide', messages = FALSE}
weekday_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by= "genres") %>%
  left_join(years_average, by = "year") %>%
  left_join(month_average, by = "month") %>%
  group_by(weekday) %>%
  summarize(day_mean = mean(rating - baseline_mean - userId_mean - movieId_mean - genre_mean - year_mean - month_mean))

ratings_prediction <- test_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by = "genres") %>%
  left_join(years_average, by = "year") %>%
  left_join(month_average, by = "month") %>%
  left_join(weekday_average, by = "weekday") %>%
  mutate(prediction = baseline_mean + userId_mean + movieId_mean + genre_mean + year_mean + month_mean + day_mean) %>%
  pull(prediction)

RMSE_user_movie_year_genre_month_day_effect <- RMSE(ratings_prediction, test_set$rating, na.rm = T)
````

````{r, echo = FALSE}
rmse_results <- rmse_results %>% rbind(c("effect of user, movie, genre, year, month and weekday", round(RMSE_user_movie_year_genre_month_day_effect, 6)))
kable(rmse_results)
````

#### **2.5 Regularization** \
\
````{r,  warning = FALSE, echo = FALSE, results = 'hide', messages = FALSE}
lambdas <- seq(0, 10, 0.5)

RMSES <- sapply(lambdas, function(l){
  userId_average <- train_set %>%
    group_by(userId) %>%
    summarize(userId_mean = sum(rating - baseline_mean)/(n()+l))
  
  movieId_average <- train_set %>%
    left_join(userId_average, by = "userId") %>%
    group_by(movieId) %>%
    summarize(movieId_mean = sum(rating - baseline_mean - userId_mean)/(n()+l))
  
  genres_average <- train_set %>%
    left_join(userId_average, by = "userId") %>%
    left_join(movieId_average, by = "movieId") %>%
    group_by(genres) %>%
    summarize(genre_mean = sum(rating - baseline_mean - userId_mean - movieId_mean)/(n()+l))
  
  years_average <- train_set %>%
    left_join(userId_average, by = "userId") %>%
    left_join(movieId_average, by = "movieId") %>%
    left_join(genres_average, by = "genres") %>%
    group_by(year) %>%
    summarize(year_mean = sum(rating - baseline_mean - userId_mean - movieId_mean - genre_mean)/(n()+l))
  
  month_average <- train_set %>%
    left_join(userId_average, by = "userId") %>%
    left_join(movieId_average, by = "movieId") %>%
    left_join(years_average, by = "year") %>%
    left_join(genres_average, by= "genres") %>%
    group_by(month) %>%
    summarize(month_mean = sum(rating - baseline_mean - userId_mean - movieId_mean - genre_mean - year_mean)/(n()+l))

  weekday_average <- train_set %>%
    left_join(userId_average, by = "userId") %>%
    left_join(movieId_average, by = "movieId") %>%
    left_join(genres_average, by= "genres") %>%
    left_join(years_average, by = "year") %>%
    left_join(month_average, by = "month") %>%
    group_by(weekday) %>%
    summarize(day_mean = sum(rating - baseline_mean - userId_mean - movieId_mean - genre_mean - year_mean - month_mean)/n()+l)
  
  ratings_prediction <- test_set %>%
    left_join(userId_average, by = "userId") %>%
    left_join(movieId_average, by = "movieId") %>%
    left_join(genres_average, by = "genres") %>%
    left_join(years_average, by = "year") %>%
    left_join(month_average, by = "month") %>%
    left_join(weekday_average, by = "weekday") %>%
    mutate(prediction = baseline_mean + userId_mean + movieId_mean + genre_mean + year_mean + month_mean + day_mean) %>%
    pull(prediction)
  return(RMSE(ratings_prediction, test_set$rating, na.rm = T))
})

lambda <- lambdas[which.min(RMSES)]
`````
Next I've tried using regularization but the result was that a lambda value of 0 gave the smallest RMSE. It made no difference to the previous RMSE.
````{r}
qplot(lambdas, RMSES) 
`````

#### **4. Results** \
````{r,  warning = FALSE, echo = FALSE, results = 'hide', message = FALSE}
userId_average <- train_set %>%
  group_by(userId) %>%
  summarize(userId_mean = sum(rating - baseline_mean)/(n()+lambda))

movieId_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  group_by(movieId) %>%
  summarize(movieId_mean = sum(rating - baseline_mean - userId_mean)/(n()+lambda))

genres_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  group_by(genres) %>%
  summarize(genre_mean = sum(rating - baseline_mean - userId_mean - movieId_mean)/(n()+lambda))

years_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by = "genres") %>%
  group_by(year) %>%
  summarize(year_mean = sum(rating - baseline_mean - userId_mean - movieId_mean - genre_mean)/(n()+lambda))

month_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(years_average, by = "year") %>%
  left_join(genres_average, by= "genres") %>%
  group_by(month) %>%
  summarize(month_mean = sum(rating - baseline_mean - userId_mean - movieId_mean - genre_mean - year_mean)/(n()+lambda))

weekday_average <- train_set %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by= "genres") %>%
  left_join(years_average, by = "year") %>%
  left_join(month_average, by = "month") %>%
  group_by(weekday) %>%
  summarize(day_mean = sum(rating - baseline_mean - userId_mean - movieId_mean - genre_mean - year_mean - month_mean)/(n()+lambda))

ratings_prediction_validation <- validation %>%
  left_join(userId_average, by = "userId") %>%
  left_join(movieId_average, by = "movieId") %>%
  left_join(genres_average, by = "genres") %>%
  left_join(years_average, by = "year") %>%
  left_join(month_average, by = "month") %>%
  left_join(weekday_average, by = "weekday") %>%
  mutate(prediction = baseline_mean + userId_mean + movieId_mean + genre_mean + year_mean + month_mean + day_mean) %>%
  pull(prediction)

RMSE_validation <- RMSE(ratings_prediction_validation, validation$rating, na.rm=T)
`````

````{r,  warning = FALSE, echo = FALSE}
rmse_results <- rmse_results %>% rbind(c("prediction on validation data set", round(RMSE_validation, 6)))
kable(rmse_results)
````


#### **4. Conclusion** \
My final RMSE on the validation data set is **0.8819228**. As predictors I used the effect of user, movie, genre, year, month and weekday. The RMSE is larger than the RMSE required to receive maximal points in the rating by the Havardx Data Science Capstone project but smaller than the initial RMSE of **1.060708** calculated by the average movie ratings. I've tried using regularization but the result was that a lambda value of 0 gave the smallest RMSE. 


